A dynamic programming algorithm to assess the reliability of inferences from network enrichment analysis

### 2018-05-04 doc/genesetdp.tex
x
Added abstract

\documentclass[a4paper,american]{lipics-v2016}
% \usepackage[USenglish]{babel}
% \usepackage{microtype}
\bibliographystyle{plainurl}
\title{A dynamic programming algorithm to assess the reliability of inferences from network enrichment analysis}
\titlerunning{Dynamic Programming for Inferences from Network Enrichment Analysis} %optional, in case that the title is too long; the running title should fit into the top page column

@@

\begin{abstract}
FIXME
\end{abstract}

@@

A limiting factor of GEA is the pathway definition databases. While \url{http://pathguide.org} list 708 databases of pathway definitions\cite{bader2006pathguide},  some critique has been voiced concerning the completeness and rigor of current pathway databases. Hence efforts have been directed to designing methods that extend the pathway definitions, by functional association networks, like STRING\cite{szklarczyk2014string}. Instead of directly examining the overlap between the query and pathway sets, one can evaluate the number of links in the functional association network that connect the query and pathway set\cite{shojaie2010network, alexeyenko2012network, glaab2012enrichnet, mccormack2013statistical, ogris2016novel, signorelli2016neat}. This is known as Network Enrichment Analysis (NEA).

FIXME: Review the mentioned methods. Particularly which null models do they use?

The quality of the inferences from NEA has previously been assessed by Monte Carlo-simulations that investigates the scores distribution when evaluating the query set using random networks which are generated using methods with varying degrees of topological conservation.

Since these methods of network randomization are not completely random, they introduce bias in the statistics that may or may not be desirable, as the set of random networks is much larger than the ones of random networks that retain a certain topological characteristic. Also, the somewhat inaccessible process used to randomize networks produces statistics that are equaly hard to interpret as the null hypothesis becomes complex. FIXME: Should this be in the dicussion?

>>>>

\documentclass[a4paper,american]{lipics-v2016}
%\usepackage[USenglish]{babel}
%\usepackage{microtype}
\bibliographystyle{plainurl}
\title{A dynamic programming algorithm to assess the reliability of inferences from network enrichment analysis}
\titlerunning{Dynamic Programming for Inferences from Network Enrichment Analysis} %optional, in case that the title is too long; the running title should fit into the top page column

@@

\begin{abstract}
A prevailing technique to infer function from findings in high-throughput experiments is gene set enrichemnet analysis, where the sets of genes are compared to predefined sets of related genes, so called pathways. As some of such pathways are known to be incomplete, algoritmic efforts have been made to complement them with information from functional association networks, using so called Network Enrichment Analysis (NEA). Traditionally significance of inferences from NEA has been asssigned using network perturbations. Here we instead have designed a dynamic programming  algorithm that can calculte the score distribution of NEA scores and makes it possible to assign unbiased mid $p$~values to inferences. We demonstrate that our method obtains a superior statistical calibration as compared to the popular NEA inference engin, NEA.

\end{abstract}

@@

A limiting factor of GEA is the pathway definition databases. While \url{http://pathguide.org} list 708 databases of pathway definitions\cite{bader2006pathguide},  some critique has been voiced concerning the completeness and rigor of current pathway databases. Hence efforts have been directed to designing methods that extend the pathway definitions, by functional association networks, like STRING\cite{szklarczyk2014string}. Instead of directly examining the overlap between the query and pathway sets, one can evaluate the number of links in the functional association network that connect the query and pathway set\cite{alexeyenko2012network, glaab2012enrichnet, mccormack2013statistical, ogris2016novel, signorelli2016neat}. This is known as Network Enrichment Analysis (NEA).

The significance of the inferences from NEA are assessed by null models that investigate the number of expected random links from the query geneset to the pathway. Previous efforts have settled for varius parametric null distributions, where the parameters are estimated by Monte Carlo-simulations that permute the network definitions by methods with varying degrees of topological conservation.

Since such methods of network randomization are not completely random, they introduce bias in the statistics that may or may not be desirable, as the set of random networks is much larger than the ones of random networks that retain a certain topological characteristic. Also, the somewhat inaccessible process used to randomize networks produces statistics that are equaly hard to interpret as the null hypothesis becomes complex. FIXME: Should this be in the dicussion?
